---
title: "Scanpy in R"
output: html_document
---

What is Scanpy?
===============

Scanpy is a scalable Python toolkit for analyzing single-cell gene expression
data. It includes functionality commonly used during an analysis of scRNA-seq
data including quality control, clustering of cells and testing for marker
genes. For more information please have a look at the
[Scanpy documentation][scanpy-docs] or read the
[associated publication][scanpy-pub].

Why would I want to use it from R?
==================================

There are various reasons you may want to interact with the Scanpy package from
R. You may already be using Scanpy for your project but want to use an R
package for a section of your analysis or your project but be R based and you
want to use some of Scanpy's functionality.

What is covered in this tutorial?
=================================

This tutorial will cover how to set up a environment that lets you interact with
Scanpy from R. It will demonstrate some of the functions in Scanpy but it won't
cover everything the package can do or how best to use Scanpy for your analysis.
For more details about what Scanpy can do please refer to the tutorials and
other information in the [Scanpy documentation][scanpy-docs].

Setting up
==========

To use Scanpy from R you will (unsurprisingly) need both an R and Python
environment set up. If you are mostly an R user the easiest approach is to use
the **{renv}** package to setup and manage these environments. This is what we
will do for this tutorial. If you are primarily a Python user you may find it
easier to create your own **conda** or **virtualenv** environment. The main
thing is to have both R and Python available (with the required packages
installed), however it is highly recommended you use some kind of environment
system to improve the reproducibilty of your project.

Using **{renv}**
----------------

The [**{renv}**][renv] package helps you reproducible environments for an R
project. These environments are designed to be isolated, portable and
reproducible. This is similar to **conda** environments but they are confined
to a specific project directory. Please refer to the
[**{renv}** documentation][renv] for more details.

### Creating the R environment

To use **{renv}** we first need to make sure it is installed:

```{r install-env, eval = FALSE}
install.packages("renv")
```

We can then ask **{renv}** to create an envronment for us:

```{r renv-init, eval = FALSE}
renv::init()
```

This command creates a new `renv/` folder where your installed packages (and
some other things) will be stored and a `renv.lock` file which will be used to
record the package versions. It will also create (or modify) the `.Rprofile`
file to make sure the environment is activated. Once we have this environment
activated we can install/remove packages  but they will be kept separate from
the global package environment. If you already have files with R code in this 
directory **{renv}** will attempt to find which packages you are using and
install them into your new library. By default `renv::init()` will create the 
environment in the current working directory but you can also tell it to create
it somewhere else by setting the `project` argument.

### Creating the Python environment

As we are going to be using Python we need to make sure we have the
[**{reticulate}**][reticulate] package installed in our new environment:

```{r intall-reticulate, eval = FALSE}
renv::install("reticulate")
```

This package provides a set of tools for interoperability between Python and R
(the Python equivalent is the [**rpy2** package][rpy2]). This includes calling
Python functions from R, translating between Python and R objects and managing
Python environments from R. For more details see the
[**{reticulate}** documentation][reticulate].

Once **{reticulate}** is installed we can tell **{renv}** that we also want to
use a Python evironment:

```{r use-python, eval = FALSE}
renv::use_python()
```

Unless you specify a Python version with the `python` argument or the
`RETICULATE_PYTHON` environment variable is set your default Python installation
will be used. By default yhe type of environment created depends on your
operating system (**virtualenv** on Linux/macOS or **conda** on Windows) but if
you have a preference this can be controlled using the `type` argument.

If you choose not to use **{renv}** a Python environment can also be created
using the `reticulate::use_python()` and 
`reticulate::use_virtualenv()`/`reticulate::use_condaenv()` functions.

Installing packages
-------------------

### R packages

Use the following command to install the R packages used during this tutorial:

```{r r-packages, eval = FALSE}
pkgs <- c(
    "renv",
    "reticulate"
)

# If you are using an {renv} environment
renv::install(pkgs)

# Otherwise do it the normal way
install.packages(pkgs)
```

### Python packages

The following Python packages are required for this tutorial:

```{r py-packages, eval = FALSE}
py_pkgs <- c(
    "scanpy"
)
```

To install them from R we can use `reticulate::py_install()`:

```{r py-pkgs-install, eval = FALSE}
reticulate::py_install(py_pkgs)
```

If you are not using **{renv}**/**{reticulate}** to manage your Python
environment please install them in the way you would normally.

[scanpy-docs]: https://scanpy.readthedocs.io/en/latest/ "Scanpy documentation"
[scanpy-pub]: https://doi.org/10.1186/s13059-017-1382-0 "Scanpy publication"
[renv]: https://rstudio.github.io/renv/ "renv documentation"
[reticulate]: https://rstudio.github.io/reticulate/ "reticulate documentation"
[rpy2]: https://rpy2.github.io/ "rpy2 website"
